Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-01-23 paper.tex
Added the 'initial measurements' section, conclusion and abstract.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\begin{abstract}
%\boldmath
The abstract goes here.

@@

This can essentially be distilled into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature. In case of non-blocking DDL statements queries issued by the application for instance may proceed, whereas with blocking DDL statements these queries may be put on hold for the duration of the schema modification. This disruption can range anywhere from milliseconds to hours depending on the DDL statement, the size of the affected table, and database implementation. One such example would be adding a new non-nullable column to a PostgreSQL database. This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 5 minutes.}
\end{itemize}

@@

% Technically we could go without this subsubsection...
\subsubsection{PRISM}
PRISM~\cite{Curino:2010:URI:1921071.1921078, Curino:2009:ADS:1656437.1656444} is a tool which is a part of a larger framework. Although PRISM does not solve the problem of downtime when performing operations on the database schema, it does aim to shorten it. PRISM is designed to reason about the database schema changes, and infer how to rewrite existing queries to the new resulting database schema. In addition it can generate the migration and rollback scripts which will transform the current database schema to the new database schema and back. The most influential concept of PRISM is the definition of several new operations to evolve a database schema. These described Schema Modification Operations (\textit{SMOs}), offer ways to split or merge tables into multiple tables (structurally or content-wise).

@@

\subsection{Road ahead}

As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.

\begin{itemize}
  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change. To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
  \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables. It does not yet support the entire PostgreSQL query syntax.}
  % Technically speaking this next one is optional.%
  \item{Adding support for SMOs. These operations allow for more complex schema operations on tables, and should enable software engineers to make more elaborate changes to the database schema.}
  \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
\end{itemize}

\section{Early evaluation} %(0.5)

\begin{itemize}
  \item{Experiment setup}
  \item{Results}
\end{itemize}

@@

\section{Conclusion} 
The conclusion goes here.

>>>>

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  \usepackage{subfig}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%

\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\begin{abstract}
%\boldmath

Continuous Deployment is a hot topic. It holds much promise for developing and deploying software applications, yet both software engineers and release engineers face various challenges when adopting this practice. In our view, the biggest technical and operational challenge is evolving the database schema of SQL databases without disrupting the clients connected to the database. It also can have an impact on how the application is developed, since big schema changes cannot be deployed without downtime, requiring intermediate versions of the application. In this paper we explore tools, and approaches to address this problem.

@@

This can essentially be distilled into two separate problems:
\begin{itemize}
  \item{Given the deployment of a new version of a web application, depending on the deployment strategy it might be required that both the old and the new version of the application run in parallel (\textit{Rolling Upgrades}, or \textit{Canary Releases}). This means that the schema on the SQL database needs to be compatible with the schema expected by both the old and the new application.}
  \item{Operations which modify the schema in SQL databases (\textit{DDL statements}), depending on the SQL database implementation may be either non-blocking, or blocking in nature. In case of non-blocking DDL statements queries issued by the application for instance may proceed, whereas with blocking DDL statements these queries may be put on hold for the duration of the schema modification. This disruption can range anywhere from milliseconds to hours depending on the DDL statement, the size of the affected table, and database implementation. One such example would be adding a new non-nullable column to a PostgreSQL database. This operation is blocking in nature --- preventing other queries from running while it is being executed --- and when applied to a table of 100 million rows (in our tests) blocked other queries from running on that table for 2.5 minutes.}
\end{itemize}

@@

@@

\section{Initial measurements} %(0.5)

We feel it is important to validate our approach at an early stage. Although QuantumDB is far from ready, we setup a small experiment to verify that the concept and our implementation of it works, and that it can match the capabilities of the tools offered in practice.

\subsection{Setup}
We have installed PostgreSQL 9.3 and a test application --- available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory. The test application inserts 100 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements. It then uses one of two methods to add a non nullable column to the ``users'' table. During this process we keep track of the performance of the queries issued by the demo application and how long each method takes to complete.

The ``naive'' approach simply executes an ``\textit{ALTER TABLE users ADD COLUMN}'' query to add the new column to the database. Once this query has completed, we can start a new demo application which can operate on the new structure of the ``users'' table. Since the new structure of the ``users'' table is compatible with both the old and the new version of the demo application they can run in parallel to each other, allowing for any kind of deployment method to be used.

Our approach creates a ghost table like much the tools from practice do. This ghost table is created with the new structure of the ``users'' table, and then filled iteratively in batches of 2,000 records at intervals of 50 milliseconds in an attempt to avoid negatively affecting the performance of the queries issued by the demo application. Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which expose each demo application to a different database schema (containing either the ``users'' table or the ghost table), while QuantumDB core keeps both tables sychronized to each other using database triggers.

\subsection{Results}

The total migration time for the ``naive'' approach was 2 minutes and 26 seconds, during which the demo application was unable to read from or write to the database. The total migration time for the QuantumDB approach was 50 minutes and 41 seconds. Although significantly slower than the ``naive'' approach, the demo applications remained operational throughout this period (see Figure~\ref{fig:migration}).

\begin{figure}%
    \centering
    \subfloat[Before migration]{{\includegraphics[width=1.55in]{normal-operations.png} }}%
    \qquad
    \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
    \caption{Query performance of demo application before and during migration}%
    \label{fig:migration}%
\end{figure}

It is worth noting, that we have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and database clients. During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough, resulting in queries taking longer to complete than regularly. We are confident that we can greatly reduce this problem by tweaking the PostgreSQL config, or switching to a faster hard disk.

\section{Road ahead}
As stated before, the implementation of QuantumDB is not yet ready for production. To be able to use QuantumDB effectively in production environments, we still need to implement some features.

\begin{itemize}
  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change. To fully support foreign keys QuantumDB would also need to be able to deal with foreign keys referring to tables under change.}
  \item{Use metrics to throttle the copy phase. When the database is (too) busy we can slow down the iterative copy process, and vice-versa speed it up when the database is not under significant load.}
  \item{QuantumDB currently only supports PostgreSQL, but should prove easily extensible to other popular SQL databases since most of these have support for triggers and procedures.}
  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables. It does not yet support the entire PostgreSQL query syntax.}
  \item{Explore if Facebook's OSC approach can also be implemented as an alternative migration algorithm, and see where both algorithms differ in performance.}
  \item{Evaluate QuantumDB more extensively in use cases at different organizations and in production environments.}
\end{itemize}

@@

\section{Conclusion}

Although we are still in the early stages of developing QuantumDB, initial measurements are promising. With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still remaining operational. Applications can connect with the database according to a specific schema version, and no longer have to deal with database schemas being in an in-between state. This effectively decouples the release process from the development process.
